---
title: 'Reproducible Research : Course Project 1'
output: html_document
---

##Loading and preprocessing the data

Loading packages *dplyr*, *lubridate* and *ggplot2*.

```{r, results="hide"}
library(dplyr)
library(lubridate)
library(ggplot2)

activity <- read.csv("activity.csv")

activity$date <- ymd(activity$date)
```

##What is mean total number of steps taken per day?

```{r}
activity_day <- (activity %>% group_by(date))

totalsteps <- summarize(activity_day, sum_steps=sum(steps, na.rm=TRUE))

qplot(sum_steps, data=totalsteps, xlab="Total number of steps per day")
```

## What is mean total number of steps taken per day?

```{r}
meansteps <- mean(totalsteps$sum_steps, na.rm=TRUE)
mediansteps <- median(totalsteps$sum_steps, na.rm=TRUE)
```

The mean is **`r meansteps`**. 
The median is **`r mediansteps`**

##What is the average daily activity pattern?

```{r}
activity_interval <- (activity %>% group_by(interval))

meaninterval <- summarize(activity_interval, mean=mean(steps, na.rm=TRUE))

qplot(interval, mean, data=meaninterval, geom="line")

maxinterval <- filter(meaninterval, mean==max(mean))$interval
```

The 5-minute interval that, on average, contains the maximum number of steps is interval **`r maxinterval`**.

##Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
NA_number <- sum(is.na(activity$steps))
```

The number of missing value is **`r NA_number`**.

##Filling the missing values for each interval with the mean for that 5-minute interval.

```{r}
activity_filled <- merge(activity, meaninterval, by="interval")

## column "fill" containing value for filling
activity_filled$fill <- is.na(activity_filled$steps)*activity_filled$mean
activity_filled$steps <- rowSums(activity_filled[,c("steps","fill")], na.rm=TRUE)

## Create a new dataset that is equal to the original dataset but with the missing data filled in.

activity_filled <- select(activity_filled, 1:3)
```

```{r}
activity_filled_day <- (activity_filled %>% group_by(date))

totalsteps_filled <- summarize(activity_filled_day, sum_steps=sum(steps))

## Histogram
qplot(sum_steps, data=totalsteps_filled, xlab="Total number of steps per day")
```

## mean and median value of the filled dataframe

```{r}
meansteps_f <- mean(totalsteps_filled$sum_steps)
mediansteps_f <- median(totalsteps_filled$sum_steps)
```

The mean is **`r meansteps_f`**
The median is **`r mediansteps_f`**
Mean and median steps are higher

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels: weekday and weekend indicating whether a given date is a weekday or weekend day.

```{r}
weekday <- function(x) {
        day <- weekdays(x)
        if (day %in% c("segunda-feira", "terça-feira", "quarta-feira",
                       "quinta-feira", "sexta-feira"))
                "weekday"
        else "weekend"
}

activity_filled$day <- sapply(activity_filled$date, weekday)
activity_filled$day <- as.factor(activity_filled$day)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this lot should look like using simulated data.

```{r}
activity_day <- (activity_filled %>% group_by(interval, day) %>% summarize(mean_steps = mean(steps)))

g <- ggplot(activity_day, aes(x= interval, y=mean_steps))
g + geom_line(aes(color=day)) + facet_wrap("day", nrow=2) + 
        labs(x = "5-minute interval", y= "Number of steps")
```

